---
title: "Regression Modeling Strategies - Chapter 12"
---

Description: develop a binary logistic regression model to describe survival patterns of Titanic passengers.
- Nonparametric regression - nonparametric tests do not rely on an assumed distribution of the data
- Multiple imputation to complete missing ages

12.1 Descriptive Statistics
```{r}
require(rms)

getHdata(titanic3)

# list of names of variables to analyze
v <- c('pclass', 'survived', 'age', 'sex', 'sibsp', 'parch') #sibsp: # of siblings/spouses aboard, parch: # of parents/children aboard
t3 <- titanic3[, v] #subset original dataset
units(t3$age) <- 'years'
latex(describe(t3), file='')
```

Truncate sibsp and parch at 3 (bc not many passengers are >3) to estimate stratified survival probabilities. About 30% of observations are missing ages.

```{r}
dd <- datadist(t3)
# describe distributions of variables to rms
options(datadist='dd')
s <- summary(survived ~ age + sex + pclass +
               cut2(sibsp,0:3) + cut2(parch,0:3), data=t3)
plot(s, main='', subtitles=FALSE)
```

There are strong effects of sex and class on survival. The effect of parch and sibsp is unclear. To clarify, show four-way empirical relationships with the fraction of surviving passengers by creating four cells for sipsp x parch combinations and creating two age groups. Suppress proportions based on < 25 passengers in a cell.

```{r}
#transform variables in t3 and assign to new data frame
tn <- transform(t3, 
                agec = ifelse(age < 21, 'child', 'adult'),
                sibsp = ifelse(sibsp == 0, 'no sib/sp', 'sib/sp'),
                parch = ifelse(parch ==0, 'no par/child', 'par/child'))

g <- function(y) if(length(y) < 25) NA else mean(y)
s <- with(tn, summarize(survived, 
                        llist(agec, sex, pclass, sibsp, parch), g)) ## ERROR - can't summarize on survived because the class is labelled
ggplot(subset(s, agec != 'NA'),
       aes(x=survived, y=pclass, shape=sex)) +
  geom_point() + facet_grid(agec ~ sibsp * parch) +
  xlab('Proportion Surviving') + ylab('Passenger Class') +
  scale_x_continuous(breaks=c(0, .5, 1))

```

12.2 Exploring Trends with Nonparametric Regression

```{r}
b <- scale_size_discrete(range=c(.1, .85))
yl <- ylab(NULL)
p1 <- ggplot(t3, aes(x=age, y=survived)) +
  histSpikeg(survived ~ age, lowess=TRUE, data=t3) +
  ylim(0,1) + yl
p2 <- ggplot(t3, aes(x=age, y=survived, color=sex)) +
  histSpikeg(survived ~ age + sex, lowess=TRUE,
             data=t3) + ylim(0,1) + yl
p3 <- ggplot(t3, aes(x=age, y=survivad, size=pclass)) +
  histSpikeg(survived ~ age + pclass, lowess=TRUE,
             data=t3) + b + ylim(0,1) + yl
p4 <- ggplot(t3, aes(x=age, y=survived, color=sex, 
                     size=pclass)) +
  histSpikeg(survived ~ age + sex + pclass, 
             lowess=TRUE, data=t3) +
  b + ylim(0,1) + yl
gridExtra::grid.arrange(p1, p2, p3, p4, ncol=2)
```

```{r}
top <- theme(legend.position='top')
p1 <- ggplot(t3, aes(x=age, y=survived, color=cut2(sibsp,
        0:2))) + stat_plsmo() + b + ylim(0,1) + yl + top +
  scale_color_discrete(name='siblings/spouses')
p2 <- ggplot(t3, aes(x=age, y=survived, color=cut2(parch,
        0:2))) + stat_plsmo() + b + ylim(0,1) + yl + top +
  scale_color_discrete(name='parents/children')
gridExtra::grid.arrange(p1, p2, ncol=2)
```


12.3 Binary Logistic Model with Casewise Deletion of Missing Values

Eliminate observations that have missing data 

```{r}
f1 <- lrm(survived ~ sex*pclass*rcs(age,5) +
            rcs(age,5)*(sibsp + parch), data=t3)
latex(anova(f1), file=' ', label='titanic-anova3',
      size='small')

f <- lrm(survived ~ (sex + pclass + rcs(age, 5))^2 +
           rcs(age,5)*sibsp, data=t3)
print(f, latex=TRUE)
```

```{r}
latex(anova (f), file=' ', label='titanic-anova2 ', size='small')

p <- Predict(f, age, sex, pclass, sibsp=0, fun=plogis)
ggplot(p)
```

```{r}
ggplot(Predict(f, sibsp, age=c(10,15,20,50), conf.int=FALSE))
```

```{r}
f <- update(f, x=TRUE, y=TRUE)
set.seed(131)
latex(validate(f, B=200), digits=2, size='Ssize ')
```

```{r}
cal <- calibrate(f, B=200)
plot(cal, subtitles = FALSE)
```

12.4 Examining Missing Data Patterns
```{r}
na.patterns <- naclus(titanic3)
require(rpart)
who.na <- rpart(is.na(age) ~ sex + pclass + survived +
                  sibsp + parch, data=titanic3, minbucket=15)
naplot(na.patterns, 'na per var')
plot(who.na, margin=.1); text(who.na)
plot(na.patterns)
```

```{r}
plot(summary(is.na(age) ~ sex + pclass + survived +
               sibsp + parch, data=t3))
```

```{r}
m <- lrm(is.na(age) ~ sex * pclass + survived + sibsp + parch, 
         data = t3)
print(m, latex=TRUE, needspace='2in')
```

```{r}
latex(anova(m), file=' ', label='titanic-anova.na')
```

12.5 Multiple Imputation

```{r}
set.seed(17)
mi <- aregImpute (~ age + sex + pclass +
                    sibsp + parch + survived, 
                  data=t3, n.impute=20, nk=4, pr=FALSE)
mi
```

```{r}
mi$imputed$age[1:10, 1:10] #print first 10 imputations
```

```{r}
plot(mi)
Ecdf(t3$age, add=TRUE, col='gray', lwd=2,
     subtitles = FALSE)
```

```{r}
f.mi <- fit.mult.impute(
  survived ~ (sex + pclass + rcs(age,5))^2 +
    rcs(age,5)*sibsp,
  lrm, mi, data=t3, pr=FALSE)
latex(anova(f.mi), file = ' ', label='titanic-anova.mi',
      size='small')
)
```
```{r}
p1 <- Predict(f, age, pclass, sex,  sibsp=0, fun=plogis)
p2 <- Predict(f.mi, age, pclass, sex,  sibsp=0, fun=plogis)
p <- rbind('Casewise Deletion'=p1, 'Multiple Imputation'=p2)
ggplot(p, groups='sex', ylab='Probability of Surviving')
```

12.6 Summarizing the Fitted Model

```{r}
s <- summary(f.mi, age=c(1,30), sibsp=0:1)
plot(s, log=TRUE, main=' ')
```

```{r}
phat <- predict(f.mi, 
                combos <- 
                  expand.grid(age=c(2, 21, 50), sex=levels(t3$sex),
                              pclass=levels(t3$pclass),
                              sibsp=0), type = 'fitted')

options(digits=1)
data.frame(combos, phat)
```

```{r}
options(digits=5)
pred.logit <- Function(f.mi)
latex(pred.logit, file='', type='Sinput', size='small',
      width.cutoff=49)
```

```{r}
plogis(pred.logit(age=c(2,21,50), sex='male', pclass='3rd'))
```

