---
title: "Regression Modeling Strategies - Chapter 11"
output:
  html_notebook: default
  pdf_document: default
---


```{r}
require(rms)
getHdata(prostate)
head(prostate)
```

11.3 Data Transformations and Single Imputation

```{r}
prostate <- 
  within(prostate, {
    levels(ekg)[levels(ekg) %in%
                  c('old MI', 'recent MI')] <- 'MI' #combine old MI and recent MI in MI
    ekg.norm <- 1*(ekg %in% c('normal','benign')) #assign value of 1 to ekg.norm if ekg is normal or benign
    levels(ekg) <- abbreviate(levels(ekg)) #abbreviate strings to minimum length characters
    pfn <- as.numeric(pf)
    levels(pf) <- levels(pf)[c(1,2,3,3)] #assign numeric values to categories in pf
    cvd <- status %in% c("dead - heart or vascular",
                         "dead - cerebrovascular") #assign value of TRUE if status is dead - heart or vascular or dead - cerebrovascular
    rxn = as.numeric(rx) })

```

```{r}
ptrans <- 
  transcan(~ sz + sg + ap + sbp + dbp +
             age + wt + hg + ekg + pf + bm + hx + dtime + rx,
           imputed=TRUE, transformed=TRUE,
           data=prostate, pl=FALSE, pr=FALSE)

imp <- impute(ptrans, data=prostate, list.out=TRUE)
```

```{r}
NAvars <- all.vars(~ sz + sg + age + wt + ekg)
for(x in NAvars) prostate [[x]] <- imp[[x]]
subset <- prostate$status %in% c("dead - heart or vascular", 
                                 "dead - cerebrovascular", "dead - prostatic ca")
trans <- ptrans$transformed[subset,]
psub <- prostate[subset,]
```

11.4 Regression on Original Variables, Principal Components and Pretransformations

How well do PCs on raw and transformed variables predict the cause of death?

```{r}
sum(prostate$cvd) #there are 127 cases where death is due to heart or vascular or cerebrovascular
```

```{r}
ipc <- function(x, k=1, ...)
  princomp(x, ..., cor=TRUE)$scores[,1:k] #function to compute the first k PCs

#compute the first 8 PCs on raw variables then transformed ones
pc8 <- ipc( ~ sz + sg + log(ap) + sbp + dbp + age +
              wt + hg + ekg.norm + pfn + bm + hx + rxn + dtime, 
            data=psub, k=8)
f8 <- lrm(cvd ~ pc8, data=psub)
pc8t <- ipc(trans, k=8)
f8t <- lrm(cvd ~ pc8t, data=psub)

#fit binary logistic model on original variables
f <- lrm(cvd ~ sz + sg + log(ap) + sbp + dbp + age +
              wt + hg + ekg + pf + bm + hx + rx + dtime, data=psub)

#expand continuous variables using splines
g <- lrm(cvd ~ rcs(sz,4) + rcs(sg,4) + rcs(log(ap),4) +
           rcs(sbp,4) + rcs(dbp,4) + rcs(age,4) + rcs(wt,4) +
           rcs(hg,4) + ekg + pf + bm + hx + rx + rcs(dtime,4),
         data=psub)

#fit binary logistic model on individual transformed var
h <- lrm(cvd ~ trans, data=psub)

#compare the five approaches to modeling using AIC  - smaller is better

c(f8=AIC(f8), f8t=AIC(f8t), f=AIC(f), g=AIC(g), h=AIC(h))
```


11.5 Description of Fitted Model
```{r}
print(f, latex=TRUE)
```

```{r}
an <- anova(f)
latex(an, file='', table.env=FALSE) #got an error message when executing this code
```

```{r}
plot(an)
s <- f$stats
gamma.hat <- (s['Model L.R.'] - s['d.f.'])/s['Model L.R.']
```

```{r}
dd <- datadist(psub); options(datadist='dd')
ggplot(Predict(f), sepdiscrete='vertical', vnames='names',
       rdata=psub,
       histSpike.opts=list(frac=function(f) .1*f/max(f) ))
```

```{r}
plot(summary(f), log=TRUE)
```

11.6 Backwards Step-Down

Identify the variables that explain the bulk of the cause of death - use fastbw() or fast backward step-down 

```{r}
fastbw(f)
```

```{r}
fred <- lrm(cvd ~ sz + log(ap) + age + hx, data=psub)
latex(fred, file='')
```

```{r}
nom <- nomogram(fred, ap=c(.1, .5, 1, 5, 10, 50),
                fun = plogis, funlabel="Probability",
                fun.at=c(.01, .05, .1, .25, .5, .75, .9, .95, .99))
plot(nom, xfrac=.45)
```

```{r}
f <- update(f, x=TRUE, y=TRUE)
v <- validate(f, B=200, bw=TRUE)
```
```{r}
latex(v, B=20, digits=3)
```

```{r}
cal <- calibrate(f, B=200, bw=TRUE)
plot(cal)
```

```{r}
vfull <- validate(f, B=200)
latex(vfull, digits=3)
```

```{r}
v5 <- validate(f, bw=TRUE, sls=0.5, type='individual', B=200)
```

```{r}
latex(v5, digits=3, B=0)
```

11.7 Model Approximation

Approximate the full model as an alternative to stepwise variable selection

```{r}
lp <- predict(f) #compute linear predictor from fill model
# insert sigma=1 as otherwise sigma=0 will cause problems
a <- ols(lp ~ sz + sg + log(ap) + sbp + dbp + age + wt +
           hg + ekg + pf + bm + hx + rx + dtime, sigma=1,
         data=psub)
# specify silly stopping criterion to remove all variables
s <- fastbw(a, aics=10000)
betas <- s$Coefficients #matrix, rows=iterations
X <- cbind(1, f$x) #design matrix
# Compute the series of approximations to lp
ap <- X %*% t(betas)
# For each approx. compute approximation R^2 and ratio of
# likelihood ratio chi-square for approximate model to that of original model
m <- ncol(ap) - 1 #all but intercept-only model
r2 <- frac <- numeric(m)
fullchisq <- f$stats['Model L.R.']
for (i in 1:m) {
  lpa <- ap[,i]
  r2[i] <- cor(lpa, lp)^2
  fapprox <- lrm(cvd ~ lpa, data=psub)
  frac[i] <- fapprox$stats['Model L.R.'] / fullchisq
} 
plot(r2, frac, type='b',
     xlab=expression(paste('Approximation ', R^2)),
     ylab=expression(paste('Fraction of ',
                           chi^2, ' Preserved')))
abline(h=.95, col=gray(.83)); abline(v=.95, col=gray(.83))
abline(a=0, b=1, col=gray(.83))
```

```{r}
fapprox <- ols(lp ~ sz + sg +log(ap) + age + ekg + pf + hx +
                 rx, data=psub)
fapprox$stats['R2']
```

```{r}
latex(fapprox, file='')
```

```{r}
nom <- nomogram(fapprox, ap=c(.1, .5, 1, 5, 10, 20, 30, 40),
                fun=plogis, funlabel="Probability",
                lp.at=(-5):4,
                fun.lp.at=qlogis(c(.01, .05, .25, .5, .75, .95, .99)))
plot(nom, xfrac=.45)
```

